Paper Analysis

### 2025-01-20 Paper_WriteUp/Paper_Plots.qmd
tidying this up a lot

1. We select 10 players from each group
and look at their first and last 1000 games.
We fit our model to each of these separately
and examine posterior predictive checks based on this.

1. Then we used the model fit to the first 1000 games
to construct posterior predictive checks for the final
1000 games and see how they compare.

## 1700-1900 Players

```{r ppc 17-19}
#| eval: false
## this the global and winner effect plots
path_id <- 1
data_path <- res_data_path[path_id]

fit_17_19_first <- readRDS(file =  here(data_path, "time",
                                        "select_users_first_bullet.RDS"))
fit_17_19_last <- readRDS(file =  here(data_path, "time",
                                        "select_users_last_bullet.RDS"))

global_winner <- as_draws_df(fit_17_19_first$draws(variables = "mu_beta")) |> 
  rename("Initial Games" = mu_beta) |> 
  bind_cols(as_draws_df(fit_17_19_last$draws(variables = "mu_beta")) |> 
              rename("Later Games" = mu_beta)) |> 
  select(`Initial Games`, `Later Games`) |> 
  pivot_longer(cols = everything()) |> 
  ggplot(aes(value, fill = name)) +
  geom_histogram(position = "identity") +
  facet_wrap(~name, ncol = 1) +
  labs(fill = element_blank(), y = element_blank(), x = element_blank()) +
  theme(legend.position = "none",
        axis.text = element_text(size = axis_text_size),
        axis.title = element_text(size = axis_title),
        legend.text = element_text(size = legend_text),
        strip.text = element_text(size = axis_title),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title = element_text(size = title_size)) +
  # labs(title = "Selected 1700-1900 Bullet Players") +
  NULL

global_winner

users <- readRDS(file = here(data_path, "time", "select_users_bullet.RDS"))
## indiv winner effects
random_effect_first <- fit_17_19_first$draws() |> as_draws_df() |>
  select(starts_with("beta[")) |>
  pivot_longer(cols = everything()) |>
  mutate(param = stringr::str_extract(name, pattern = "\\d"),
         id = stringr::str_extract(name, pattern = "\\d+]"),
         id = stringr::str_replace(id, "\\]", ""),
         player_id = paste0("beta[", id, "]")) |> 
  mutate(when = "Initial")

random_effect_last <- fit_17_19_last$draws() |> as_draws_df() |>
  select(starts_with("beta[")) |>
  pivot_longer(cols = everything()) |>
  mutate(param = stringr::str_extract(name, pattern = "\\d"),
         id = stringr::str_extract(name, pattern = "\\d+]"),
         id = stringr::str_replace(id, "\\]", ""),
         player_id = paste0("beta[", id, "]")) |> 
  mutate(when = "Later")

betas <- bind_rows(random_effect_first, random_effect_last) |> 
  select(value, param, id, when)

player_id <- tibble(id = as.character(1:10), player = users)

betas |> left_join(player_id, by = "id") |> 
  mutate(player = factor(player, levels = users))

winner_effect_plot <- betas |> 
  left_join(player_id, by = "id") |> 
  mutate(player = factor(player, levels = users)) |> 
  filter(param == 2) |> 
  mutate(id = factor(id, levels = c(1:10))) |> 
  ggplot(aes(value, fill = when)) +
  geom_histogram() +
  geom_vline(xintercept = 0, alpha = 0.5) +
  scale_x_continuous(breaks = c(-0.25, 0.25)) +
  # xlim(c(-1.5, 1.5)) +
  facet_grid(rows = vars(when), cols = vars(player), scales = "free_x") +
  labs(fill = element_blank(), y = element_blank(), 
       # title = "Individual Winner Effects, Selected 1700-1900 Bullet Players",
       x = element_blank()) +
  theme_10col()
  # theme(axis.ticks.y = element_blank(),
  #       axis.text.y = element_blank(),
  #       axis.text = element_text(size = axis_text_size - 4),
  #       axis.title = element_text(size = axis_title),
  #       legend.text = element_text(size = legend_text),
  #       strip.text = element_text(size = axis_title - 6),
  #       legend.position = "none",
  #       plot.title = element_text(size = title_size))

winner_effect_plot
```

```{r ppc1 for 17-19}
#| eval: false
path_id <- 1
data_path <- res_data_path[path_id]

fit_17_19_first <- readRDS(file =  here(data_path, "time",
                                        "select_users_first_bullet.RDS"))
fit_17_19_last <- readRDS(file =  here(data_path, "time",
                                        "select_users_last_bullet.RDS"))

fit_samples <- as_draws_df(fit_17_19_first$draws())

fit_samp <- fit_samples |> 
  select(!starts_with(c("log_lik", "yrep")))

y_rep <- fit_samples |> select(starts_with("y_rep"))

rm(fit_samples)
rm(fit_samp)

stan_data_ave_first <- readRDS(file = here(data_path, "time",
                                           "stan_data_first.RDS"))

y_rep_mod <- y_rep |> 
  rownames_to_column() |>  
  pivot_longer(-rowname) |> 
  pivot_wider(names_from=rowname, values_from=value) |> 
  ## remove the draws and chains here
  mutate(focal_id = stan_data_ave_first$id)

games_won <- y_rep_mod |> 
  pivot_longer(cols = `1`:`4000`, names_to = "draw", values_to = "y") |> 
  group_by(focal_id, draw) |> 
  summarise(games_won = sum(y))

orig_data <- tibble(outcome = stan_data_ave_first$y,
                    focal_id = stan_data_ave_first$id)

orig_games_won <- orig_data |> 
  group_by(focal_id) |> 
  summarise(games_won = sum(outcome)) |> 
  mutate(focal_id = as.character(focal_id)) |> 
  left_join(player_id, by = c("focal_id" = "id"))

p1 <- games_won |> 
  mutate(focal_id = as.character(focal_id)) |> 
  left_join(player_id, by = c("focal_id" = "id")) |> 
  ggplot(aes(x = games_won)) +
  geom_histogram() +
  facet_wrap(~player, scales = "free", ncol = 5) +
  geom_vline(data = orig_games_won, 
             mapping = aes(xintercept = games_won), col = "red") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  # labs(title = "Posterior Predictive Distribution, Initial Games",
  labs(y = element_blank(),
       x = "Number of Games Won") +
  theme_5col() +
  theme(axis.title.x = element_text(size = 10,
                                    margin = margin(t = 10, b = 10)))

p1

ggsave(plot = p1, 
       filename = paste0(here("Paper_WriteUp", "paper_figures",
                              "1700_1900_ppc_1.png")), dpi = 600,
       height = 5, width = 7)

```{r ppc2 for 17-19}
#| eval: false
## time varying ppc for 1700-1900 players
draws <- fit_17_19_first$draws() |> as_draws_df() |>
  select(starts_with(c("beta[", "gamma"))) |>
  mutate(draw = row_number()) |> 
  pivot_longer(cols = "beta[1,1]":"beta[2,10]") |> 
  mutate(param = stringr::str_extract(name, pattern = "\\d"),
         id = stringr::str_extract(name, pattern = "\\d+]"),
         id = stringr::str_replace(id, "\\]", ""),
         player_id = paste0("beta[", id, "]")) |> 
  select(-name) |> 
  pivot_wider(names_from = param, values_from = value) |> 
  rename(beta1 = `1`, beta2 = `2`)

stan_data_ave_last <- readRDS(file = here(data_path, "time",
                                          "stan_data_last.RDS"))

future_games <- tibble(focal = as.character(stan_data_ave_last$id), 
       color = stan_data_ave_last$colour, 
       elo_diff = stan_data_ave_last$elo,
       hist = stan_data_ave_last$win_prop)

ppc_data <- readRDS(file = here(data_path, "time",
                                "ppc_fit_first_last.RDS"))

future_results <- tibble(focal = as.character(stan_data_ave_last$id),
                         result = stan_data_ave_last$y) |> 
  group_by(focal) |>
  summarise(total = sum(result)) |> 
  rename(id = focal) |> 
  left_join(player_id, by = "id")

last_ppc <- ppc_data |>  
  left_join(player_id, by = "id") |> 
  ggplot(aes(x = games_won)) + 
  geom_histogram() + 
  geom_vline(data = future_results, 
             mapping = aes(xintercept = total), col = "red") +
  facet_wrap(~player, scales = "free_x", ncol = 5) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  labs(y = element_blank(), x = "Number of Games Won") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  theme_5col() +
  theme(axis.title.x = element_text(size = 10,
                                    margin = margin(t = 10, 
                                                    b = 10,
                                                    unit = "pt")))

last_ppc

ggsave(plot = last_ppc, 
       filename = paste0(here("Paper_WriteUp", "paper_figures",
                              "1700_1900_ppc_2.png")), dpi = 600,
       height = 5, width = 7)
```

@@

perm_17_19 <- perm_files[1:1000] |> 
  map_dfr(~read_rds(here(perm_data_path, .x)))

@@

@@

perm_17_19_global <- sim_pars |> 
  filter(variable %in% c("mu_beta",  "gamma1", "gamma2",
                           "sigma_1", "tau[1]", "tau[2]",
                           "sigma_g1", "sigma_g2")) |> 
  ggplot(aes(mean)) +
  geom_histogram() +
  facet_wrap(~variable, scales = "free") +
  geom_vline(data = true_values,
             mapping = aes(xintercept = mean),
             col = "red") +
  theme_3col() +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  labs(x = element_blank(), y = element_blank())

@@

ggsave(plot = perm_17_19_global, 
       filename = paste0(here("Paper_WriteUp", "paper_figures",
                              "perm_17_19_global.png")), dpi = 600,
       height = 5, width = 7)
```

```{r perm winner effects}
#| eval: false
## get the selected users for this, top 20 players
path_id <- 1  ## for 1700-1900
data_path <- all_data_path[path_id]
files <- list.files(data_path)

@@

perm_17_19_winner <- player_pars |> 
  filter(param == 2) |> 
  ggplot(aes(mean)) +
  geom_histogram() +
  facet_wrap(~player_name, ncol = 5, scales = "free") +
  labs(x = element_blank(), y = element_blank()) +
  geom_vline(data = true_beta |> filter(param == 2), aes(xintercept = mean),
             col = "red") +
  theme_5col()

perm_17_19_winner

@@

```{r ppc1 and ppc2 for gm}
#| eval: false
## just do ppc plots for now
path_id <- 4
data_path <- res_data_path[path_id]
fit_gm_first <- readRDS(file =  here(data_path, "time",
                                        "select_users_first_bullet.RDS"))
fit_gm_last <- readRDS(file =  here(data_path, "time",
                                        "select_users_last_bullet.RDS"))

fit_samples <- as_draws_df(fit_gm_first$draws())

fit_samp <- fit_samples |> 
  select(!starts_with(c("log_lik", "yrep")))

y_rep <- fit_samples |> select(starts_with("y_rep"))

rm(fit_samples)
rm(fit_samp)

stan_data_ave_first <- readRDS(file = here(data_path, "time",
                                           "stan_data_first.RDS"))

y_rep_mod <- y_rep |> 
  rownames_to_column() |>  
  pivot_longer(-rowname) |> 
  pivot_wider(names_from=rowname, values_from=value) |> 
  ## remove the draws and chains here
  mutate(focal_id = stan_data_ave_first$id)

games_won <- y_rep_mod |> 
  pivot_longer(cols = `1`:`4000`, names_to = "draw", values_to = "y") |> 
  group_by(focal_id, draw) |> 
  summarise(games_won = sum(y))

orig_data <- tibble(outcome = stan_data_ave_first$y,
                    focal_id = stan_data_ave_first$id)

users <- readRDS(file = here(data_path, "time", "select_users_bullet.RDS"))

## need to give these the correct names here
player_id <- tibble(id = as.character(1:10), player = users)

orig_games_won <- orig_data |> 
  group_by(focal_id) |> 
  summarise(games_won = sum(outcome)) |> 
  mutate(focal_id = as.character(focal_id)) |> 
  left_join(player_id, by = c("focal_id" = "id"))

p1 <- games_won |> 
  mutate(focal_id = as.character(focal_id)) |> 
  left_join(player_id, by = c("focal_id" = "id")) |> 
  ggplot(aes(x = games_won)) +
  geom_histogram() +
  facet_wrap(~player, scales = "free", ncol = 5) +
  geom_vline(data = orig_games_won, 
             mapping = aes(xintercept = games_won), col = "red") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  # labs(title = "Posterior Predictive Distribution, Initial Games",
  labs(y = element_blank(),
       x = "Number of Games Won") +
  theme_5col() +
  theme(axis.title.x = element_text(size = 10,
                                   margin = margin(t = 10, b = 10, unit = "pt")),
        panel.spacing.x = unit(0.7, "lines"))

p1

ggsave(plot = p1, 
       filename = paste0(here("Paper_WriteUp", "paper_figures",
                              "gm_ppc_1.png")), dpi = 600,
       height = 5, width = 7)

draws <- fit_gm_first$draws() |> as_draws_df() |>
  select(starts_with(c("beta[", "gamma"))) |>
  mutate(draw = row_number()) |> 
  pivot_longer(cols = "beta[1,1]":"beta[2,10]") |> 
  mutate(param = stringr::str_extract(name, pattern = "\\d"),
         id = stringr::str_extract(name, pattern = "\\d+]"),
         id = stringr::str_replace(id, "\\]", ""),
         player_id = paste0("beta[", id, "]")) |> 
  select(-name) |> 
  pivot_wider(names_from = param, values_from = value) |> 
  rename(beta1 = `1`, beta2 = `2`)

stan_data_ave_last <- readRDS(file = here(data_path, "time",
                                          "stan_data_last.RDS"))

future_games <- tibble(focal = as.character(stan_data_ave_last$id), 
       color = stan_data_ave_last$colour, 
       elo_diff = stan_data_ave_last$elo,
       hist = stan_data_ave_last$win_prop)

ppc_data <- readRDS(file = here(data_path, "time",
                                "ppc_fit_first_last.RDS"))

future_results <- tibble(focal = as.character(stan_data_ave_last$id),
                         result = stan_data_ave_last$y) |> 
  group_by(focal) |>
  summarise(total = sum(result)) |> 
  rename(id = focal) |> 
  left_join(player_id, by = "id")

last_ppc <- ppc_data |>  
  left_join(player_id, by = "id") |> 
  ggplot(aes(x = games_won)) + 
  geom_histogram() + 
  geom_vline(data = future_results, 
             mapping = aes(xintercept = total), col = "red") +
  facet_wrap(~player, scales = "free_x", ncol = 5) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  labs(y = element_blank(), x = "Number of Games Won") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  theme_5col() +
  theme(axis.title.x = element_text(size = 10,
                                   margin = margin(t = 10,
                                                   b = 10,
                                                   unit = "pt")))

last_ppc

ggsave(plot = last_ppc, 
       filename = paste0(here("Paper_WriteUp", "paper_figures",
                              "gm_ppc_2.png")), dpi = 600,
       height = 5, width = 7)
```

@@

```{r}
#| eval: false
perm_data_path <- here("results/lichessGrandmasters/perm/")

@@

perm_gm_winner <- player_pars |> 
  filter(param == 2) |> 
  ggplot(aes(mean)) +
  geom_histogram() +
  facet_wrap(~player_name, ncol = 5, scales = "free") +
  labs(x = element_blank(), y = element_blank()) +
  geom_vline(data = true_beta |> filter(param == 2), aes(xintercept = mean),
             col = "red") +
  theme_5col() +
  scale_x_continuous(breaks = breaks_pretty(n = 3))

@@

perm_gm_player <- player_pars |> 
  filter(param == 1) |> 
  ggplot(aes(mean)) +
  geom_histogram() +
  facet_wrap(~player_name, ncol = 5, scales = "free") +
  labs(x = element_blank(), y = element_blank()) +
  geom_vline(data = true_beta |> filter(param == 2), aes(xintercept = mean),
             col = "red") +
  theme_5col() +
  scale_x_continuous(breaks = breaks_pretty(n = 3))

@@

## should we make this 12 players so it looks a bit better?
```

>>>>

@@

perm_17_19 <- perm_files[1:1000] |>
  map_dfr(~ read_rds(here(perm_data_path, .x)))

@@

fit_17_19_bullet_prev <- readRDS(file = paste0(
  data_path,
  "/all_rated_bullet_model_prev.RDS"
))

bullet_17_19 <- readRDS(file = paste0(
  data_path,
  "/users_bullet.RDS"
))

@@

perm_17_19_global <- sim_pars |>
  filter(variable %in% c(
    "mu_beta", "gamma1", "gamma2")) |>
  ggplot(aes(mean)) +
  geom_histogram() +
  facet_wrap(~variable, scales = "free") +
  geom_vline(
    data = true_values,
    mapping = aes(xintercept = mean),
    col = "red"
  ) +
  theme_3col() +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  labs(x = element_blank(), y = element_blank())

@@

ggsave(
  plot = perm_17_19_global,
  filename = paste0(here(
    "Paper_WriteUp", "paper_figures",
    "perm_17_19_global.png"
  )), dpi = 600,
  height = 5, width = 7
)
```

```{r figure 8 left - perm winner effects}
## get the selected users for this, top 20 players
path_id <- 1  ## for 1700-1900
data_path <- all_data_path[path_id]
files <- list.files(data_path)

@@

@@

@@

```{r figure 8 right}
perm_data_path <- here("results/lichessGrandmasters/perm/")

@@

@@

levels <- true_beta |> 
  filter(param == 1) |> 
  arrange(mean) |> 
  pull(player_name)

true_vals <- true_beta |> 
  filter(param == 1) |>  
  mutate(y = row_number() - 0.5,
         yend = y + 1)

perm_gm_player <- player_pars |> 
  filter(param == 1) |>
  mutate(y = factor(player_name, levels = levels)) |> 
  ggplot(aes(x = mean, y = y)) +
  stat_histinterval(slab_fill = NA, .width = c(0.5, 0.95)) +
  geom_segment(data = true_vals, 
               aes(x = mean, y = y, yend = yend), col = "red") +
  theme_single_y() +
  labs(x = "Estimated Player Effect", y = "")

@@

## Posterior Predictive Plots

```{r}
